---
title: "Penguins"
subtitle: Statistics, Hypothesis testing, and correlation matrix
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

#### Outcomes covered in the below work 

*For Statistics and Hypothesis Testing* 

From Working with Data (J4Y6 35):

* 4.1 Exploratory data analysis and targeted data analysis (hypothesis testing)
* 4.3 Probability and probability distributions
* 4.4 Summary statistics including measures of centrality, spread and skewness

*For Correlation Matrix*

* 4.5 Concept of correlation and correlation matrices
* 4.8 Documenting analyses

<hr>

```{r}
library(tidyverse)
library(palmerpenguins)
library(ggplot2)
library(e1071)
library(infer)
library(GGally)
```

We are interested in how bill length differ between the three species of penguin 
(Adelie, Chinstrap & Gentoo) in this dataset.

```{r}
# install.packages("palmerpenguins")
penguins <- palmerpenguins::penguins 
glimpse(penguins)
```


## 1. Exploratory analysis

We are interested in comparing bill length among species:
```{r}
penguin_species_grouped <- penguins %>%
  drop_na(bill_length_mm) %>% 
  group_by(species)
penguin_species_grouped %>% 
  summarise(count = n())
# Adelie: 151   Chinstrap: 68   Gentoo: 123	
```

### 4.1 Exploratory data analysis and targeted data analysis (hypothesis testing)

Comparing and visualising differences in each species mean bill length:
```{r}
species_mean_bill_length <- penguin_species_grouped %>% 
  select(species, bill_length_mm) %>% 
  summarise(mean_bill_length = round(mean(bill_length_mm))) 
# Adelie: 39 (38.79139)    Chinstrap: 49 (48.83382)   Gentoo: 48 (47.50488)

species_mean_bill_length %>% 
  ggplot()+
  aes(x= species, y = mean_bill_length, fill = species) +
  geom_col(stat = "identity")+
  geom_text(aes(label = mean_bill_length), vjust = -0.25) +
  labs(
    x = "Species", 
    y = "Mean Bill Length (mm)", 
    title = "Comparison of mean bill lengths among penguin species", 
    fill = "Species"
  )
```
The comparison of species bill length means shows that although Chinstrap and 
Gentoo penguins have similar average bill lengths (49mm and 48mm respectively), 
the Adelie penguin's have a lower average bill length of around 39mm. 


### 4.1 Exploratory data analysis and targeted data analysis (hypothesis testing)

However, to better understand each species, we can visualise with a histogram 
and boxplot for bill length for each species:
```{r}
# Quartiles and interquartile range: 
penguin_species_grouped %>% 
  summarise(
    Q1 = quantile(bill_length_mm, 0.25),
    Q2 = quantile(bill_length_mm, 0.5),
    Q3 = quantile(bill_length_mm, 0.75),
    IQR = Q3 - Q1
  )

# Boxplot of all bill lengths by species:
penguin_species_grouped %>% 
  ggplot() +
  aes(x = bill_length_mm, y = species, fill = species) +
  geom_boxplot() + 
  labs(
    x = "Bill lengths (mm)", 
    y = "Species", 
    title = "Comparison of bill lengths amongst species of penguin",
    fill = "Species"
  ) 

# Histogram showing number of birds with each bill length per species 
penguin_species_grouped %>%
  ggplot(aes(x = bill_length_mm, fill = species)) +
  geom_histogram(col = "white", bins = 30) +
  facet_wrap(~ species) + 
  labs(x = "Bill length (mm)", 
       y = "Number of birds with each length", 
       title = "Number of penguins with each recorded bill length", 
       fill = "Species")
```
The boxplot of all bill lenghts amongst species of penguin shows a similar story 
to the analysis of the mean bill lengths. Chinstrap and Gentoo have similar 
median bill lengths (49.6mm and 47.3mm respectfully), and Adelie penguins have a 
lower median length (38.8mm).

They all have quite a similar interquartile ranges of between 4 and 5mm (50% of 
all the penguins bill lengths in each species fall within 4 or 5mm of each 
other).

Though the histograms show a *fairly* normal distribution, the arguement for a 
slight bimodal distribution could be made especially with Chinstrap penguins. 
Further, Gentoos appear to be slightly right skewed, probably the result of an 
outlier.

It may be important to note that there were far more Adelie (n = 151) and Gentoo 
(n = 123) penguins recorded than Chinstrap penguins (n = 68).




## 2. Summary statistics

### 4.4 Summary statistics including measures of centrality, spread and skewness

```{r}
# Measures of centrality (mean and median) and spread (standard deviation)
penguin_species_grouped %>% 
  summarise(mean_bill_length = mean(bill_length_mm), 
            median_bill_length = median(bill_length_mm), 
            standard_deviation_bill_length = sd(bill_length_mm))
```

```{r}
#  Skewness using the e1071 package
#library(e1071)
penguin_species_grouped %>%
  summarise(skewness = skewness(bill_length_mm, type = 1))
# Adelie: 0.16    Chinstrap: -0.09    Gentoo: 0.64
```

The comparison of species bill length measures of centrality shows that 
Chinstrap and Gentoo penguins have similar mean and median values. Chinstrap 
bills have an average length of 48.83mm with their median recorded as 49.55mm,
and Gentoo bill lengths are 47.50mm on average with a median value of 47.30mm. 
Adelie penguin average bill lengths are shorter, recorded as 38.79mm, with a 
median value of 38.80mm. 

Each species show slightly different standard deviations. Chinstrap penguin
records differ most, on average, from their mean value with SD = 3.34. Gentoos 
standard deviation is SD = 3.08, and Adelie penguin bill lengths, on average, 
differ least from their mean value with a SD = 2.66. 

Both Adelie (0.16) and Gentoo (0.64) penguins show a positive (right) skew, and 
Chinstrap penguins show a very slight negative (left) skew (-0.08). As none of 
these values are greater than 1, this indicates that the distributions are not
highly skewed. 





## 3. Hypothesis test & probability

Perform a suitable hypothesis test to check whether the mean bill length of 
Gentoo is significantly longer than the mean bill length of Adelie.

### 4.1 Exploratory data analysis and targeted data analysis (hypothesis testing)

### -AND-
### 4.3 Probability and probability distributions

```{r}
#library(infer)
```

With a chosen significance level α = 0.05, if the final p-value is less than or
equal to 0.05 (p ≤ 0.05), this will suggest that there is a 5% probability or 
less that the observed differences were the result of sampling error (chance).

We can conduct a one-tailed, independent samples test with the following 
hypotheses:


* H0: The mean bill length of the Gentoo penguin is not significantly longer 
than that of the Adelie penguin. 

$$ H0: μ_{bill\ length\ (Gentoo)} - μ_{bill\ length\ (Adelie)} <= 0$$


* H1: The mean bill length of the Gentoo penguin is significantly longer than the 
mean bill length of the Adeli penguin.

$$ H0: μ_{bill\ length\ (Gentoo)} - μ_{bill\ length\ (Adelie)} > 0$$


Visualising all bill lengths:
```{r}
# Filtering original dataset for the species we are interested in
gentoo_and_adelie <- penguins %>% 
  drop_na(bill_length_mm) %>% 
  filter(species %in% c("Gentoo", "Adelie"))
# Visualising both species' bill length with a boxplot:
gentoo_and_adelie %>% 
  ggplot(aes(x = bill_length_mm, y = species, fill = species)) +
  geom_boxplot() +
  labs(
    x = "Bill length (mm)", 
    y = "Species", 
    title = "Comparison of bill lengths between penguin species", 
    fill = "Species")
```
The overlap here appears to mostly be between the lower 25% of the Gentoo 
penguins and the upper 25% of the Adelie penguins. 


To see if there is an implication of species upon bill length or whether the 
difference is down to sampling variation (chance), we can start by generating
the null distribution by permutation:

* Specify: it is the relationship between the bill length and the species we 
  are testing. 
  
* Hypothesise: the null hypothesis is that there is no relationship, i.e. 
  they are independent. 
  
* Generate: under the H0, there is no relationship between the two species and 
  their bill lengths. So, using permutation, the "species" variable labels are 
  shuffled (i.e.detaching the labels from rows and then randomly assigning them 
  back to rows) so that any relationship between species and bill length is 
  lost. In this case, this is repeated 5000 times to account for sampling 
  variability so generates a distribution of 5000 samples. 

* Calculate: checking the difference in the means. As the sample statistic is
  the mean of Gentoo bill lengths minus the mean of Adelie bill lengths, so this 
  is the order we specify them in in the calculation step. 
  
```{r}
null_distribution_g_and_a <- gentoo_and_adelie %>% 
  specify(bill_length_mm ~ species) %>%  
  hypothesise(null = "independence") %>% 
  generate(reps = 5000, type = "permute") %>% 
  calculate(stat = "diff in means", order = c("Gentoo", "Adelie"))
null_distribution_g_and_a
```

<br>
Calculating the observed statistic and visualising the null distribution.
* "Direction = "right"" is chosen due to this being a one-tailed test.
* On the graph, the red line indicates where the observed statistic lies on the 
distribution: 
```{r}
observed_stat_g_and_a <- gentoo_and_adelie %>% 
  specify(bill_length_mm ~ species) %>%
  calculate(stat = "diff in means", order = c("Gentoo", "Adelie"))
observed_stat_g_and_a
# 8.71
null_distribution_g_and_a %>% 
  visualise()+
  shade_p_value(obs_stat = observed_stat_g_and_a, direction = "right")
```

From the visualisation, we can see that the observed statistic (red line) is far 
to the right of our null distribution. There would therefore appear to be a very 
small probability of getting a more extreme value than our observed statistic. 
To be sure, the p-value should be calculated:

```{r}
p_value_g_and_a <- null_distribution_g_and_a %>% 
  get_p_value(obs_stat = observed_stat_g_and_a, direction = "right")
p_value_g_and_a
# p = (close to) 0 
format.pval(p_value_g_and_a)
# p < 2.22e-16
```
If p ≤ 0.05 (the initially chosen significance level), this will suggest that 
there is a 5% probability or less that the observed differences in mean lengths 
of the two species of penguin's bills were the result of sampling error (random 
chance).

Here, p is close to 0 (p < 2.22e-16). 
As p < 0.05 we are therefore able to reject the null hypothesis in favour of the 
alternative hypothesis. 
We can conclude that there is enough evidence within the data to suggest that 
Gentoo penguin's bill lengths are statistically significantly longer than those 
of the Adelie penguins. 

<hr>



## Correlation Matrix

#### Outcomes covered by the below:

* 4.5 Concept of correlation and correlation matrices
* 4.8 Documenting analyses

### Correlation matrix

A correlation matrix that shows the correlation between bill length, bill depth, 
flipper length, and body mass across species:

```{r}
#library(GGally)
```

Remember there are 2 that have NA under all records: 
(Total of each = Adelie: 151   Chinstrap: 68   Gentoo: 123)

```{r}
penguins_correlation <- penguins %>% 
  select(species, 
         bill_length_mm, 
         bill_depth_mm, 
         flipper_length_mm, 
         body_mass_g) %>% 
  drop_na()
```

```{r}
ggcorr(penguins_correlation, 
       method = c("everything", "pearson"), 
       hjust = 0.75, 
       size = 4, 
       layout.exp = 1,
       legend.size = 10, 
       label = TRUE, 
       label_size = 4) +
  labs(title = "Correlations across penguin species") 

```
It is first important to note that the data should be visualised before trusting 
the correlation coefficients above. 

The correlation matrix above shows that body mass and flipper length are the 
most strongly correlated of all the relationships. As we can see a *very strong* 
positive correlation (r = 0.9), it may be very likely that penguins with a 
larger mass have longer flipper lengths. Further *strong*, positive, correlations 
can be seen between body mass and bill length (r = 0.6) and between flipper 
length and bill length (r = 0.7). 

Negative correlations can also be seen here. There is a *weak* negative correlation 
between bill depth and bill length (-0.2) meaning that as one decreases in size 
the other may increase. A *moderate* negative correlation can be seen 
between body mass and bill depth (r = -0.5) again meaning that as one decreases 
the other may increase. Finally, a *strong* negative correlation can be 
seen between flipper length and bill depth (r = -0.6) meaning that penguins with 
longer flippers may be more likely to have shorter bill depths and vice versa. 

Therefore, (to take one example) it may be the case that the larger the mass 
of a penguin, the longer the flipper length and bill length, but the shorter the 
bill depth.  


